---
title: Data Visualisation with ggplot2
teaching: 80
exercises: 35
source: Rmd
---

```{r setup, include=FALSE}
source("data/download_data.R")
```

:::: instructor

- This episode is a broad overview of ggplot2 and focuses on (1) getting
  familiar with the layering system of ggplot2, (2) using the argument `group`
  in the `aes()` function, (3) basic customization of the plots.
- The episode depends on data created in the Data Wrangling with tidyr
  episode. If you did not get to or through all of the tidyr episode,
  you can have the learners access the data by either downloading it or
  quickly creating it using the tidyr code below. You will probably want to
  copy the code into the Etherpad.
- If you did skip the tidyr episode, you might want to go over the exporting
  data section in that episode.

::::::::::::

::::::::::::::::::::::::::::::::::::::: objectives

- Produce scatter plots, boxplots, and barplots using ggplot.
- Set universal plot settings.
- Describe what faceting is and apply faceting in ggplot.
- Modify the aesthetics of an existing ggplot plot (including axis labels and colour).
- Build complex and customized plots from data in a data frame.
- Recognize the differences between base R, lattice, and ggplot visualizations.

::::::::::::::::::::::::::::::::::::::::::::::::::

:::::::::::::::::::::::::::::::::::::::: questions

- What are the components of a ggplot?
- What are the main differences between R base plots, lattice, and ggplot?
- How do I create scatterplots, boxplots, and barplots?
- How can I change the aesthetics (ex. colour, transparency) of my plot?
- How can I create multiple plots at once?

::::::::::::::::::::::::::::::::::::::::::::::::::

* load ggplot

```{r load-package, message=FALSE, purl=FALSE}
library(tidyverse)
library(ggplot2)
```

If not still in the workspace, load the data we saved in the previous lesson.

```{r load-data, purl=FALSE}
interviews_plotting <- read_csv("data_output/interviews_plotting.csv")
```

If you were unable to complete the previous lesson or did not save the data,
then you can create it now. Either download it using `read_csv()` (Option 1)
or create it with the **dplyr** and **tidyr** code (Option 2).

### Download the data

```{r, purl=FALSE, eval=FALSE}
interviews_plotting <- read_csv("https://raw.githubusercontent.com/datacarpentry/r-socialsci/main/episodes/data/interviews_plotting.csv")
```


## Visualization Options in R

* Several options 
* Base R quick and simple 
* Simple scatterplot in base R using the `no_membrs` and `liv_count` variables:

```{r, purl=FALSE}
plot(interviews_plotting$no_membrs, interviews_plotting$liv_count,
     main = "Base R Scatterplot",
     xlab = "Number of Household Members",
     ylab = "Number of Livestock Owned")
```

### **`Lattice`**

* Lattice 


## Plotting with **`ggplot2`**

* **`ggplot2`** easy to tell what variables to plot, how they are displayed, and general visual
properties. 
* we only need minimal changes if the underlying data
change or if we decide to change from a bar plot to a scatterplot.

* **`ggplot2`** 'long' format

* layers of elements = flexi and custom
* need Data, 
* needAesthetic mapping (aes), Describes how variables are mapped onto graphical attributes
* Visual attribute of data including x-y axes, color, fill, shape, and alpha
* Geometric objects (geom): graphics rendering of objects 
* use the `ggplot()` function and bind the plot to a specific data frame.

```{r ggplot-steps-1, eval=FALSE, purl=FALSE}
interviews_plotting %>%
    ggplot()
```

- define a mapping (using the aesthetic (`aes`) function), x,y,color...

```{r ggplot-steps-2, eval=FALSE, purl=FALSE}
interviews_plotting %>%
    ggplot(aes(x = no_membrs, y = number_items))
```

- add 'geoms' – graphical representations of the data in the plot (points,
  lines, bars). **`ggplot2`** offers many different geoms; we will use some
  common ones today, including:
  
  - `geom_point()` for scatter plots, dot plots, etc.
  - `geom_boxplot()` for, well, boxplots!
  - `geom_line()` for trend lines, time series, etc.

To add a geom to the plot use the `+` operator. Because we have two continuous variables, let's use `geom_point()` first:

```{r first-ggplot, purl=FALSE}
interviews_plotting %>%
    ggplot(aes(x = no_membrs, y = number_items)) +
    geom_point()
```

The `+` in the **`ggplot2`** package is particularly useful because it allows
you to modify existing `ggplot` objects. End of line 

```{r first-ggplot-with-plus, fig.alt="Scatter plot of number of items owned versus number of household members.", eval=FALSE, purl=FALSE}
# Assign plot to a variable
interviews_plot <- interviews_plotting %>%
    ggplot(aes(x = no_membrs, y = number_items))

# Draw the plot as a dot plot
interviews_plot +
    geom_point()
```

* Example of several building blocks

```{r ggplot-with-plus-position, eval=FALSE, purl=FALSE}
## This is the correct syntax for adding layers
interviews_plot +
    geom_point()

## This will not add the new layer and will return an error message
interviews_plot
+ geom_point()
```

## Building your plots iteratively

```{r create-ggplot-object, fig.alt="Scatter plot of number of items owned versus number of household members.", purl=FALSE}
interviews_plotting %>%
    ggplot(aes(x = no_membrs, y = number_items)) +
    geom_point()
```

Fewer points than expected. This should lead us to believe that there may be multiple
observations plotted on top of each other (e.g. three observations where
`no_membrs` is 3 and `number_items` is 1).

There are two main ways to alleviate overplotting issues:

1. changing the transparency of the points
2. jittering the location of the points

Let's first explore option 1, changing the transparency of the points. What we
mean when we say "transparency" we mean the opacity of point, or your ability to
see through the point.

```{r adding-transparency, fig.alt="Scatter plot of number of items owned versus number of household members, with transparency added to points.", purl=FALSE}
interviews_plotting %>%
    ggplot(aes(x = no_membrs, y = number_items)) +
    geom_point(alpha = 0.5)
```

* Help a little only. Jittering introduces a little bit
of randomness into the position of our points. 
* Note that this solution is suitable for plotting 
integer figures, while for numeric figures with decimals, geom_jitter() becomes 
inappropriate because it obscures the true value of the observation.

We can jitter our points using the `geom_jitter()` function instead of the
`geom_point()`  function, as seen below:

```{r adding-jitter, fig.alt="Scatter plot of number of items owned versus number of household members, showing jitter.", purl=FALSE}
interviews_plotting %>%
    ggplot(aes(x = no_membrs, y = number_items)) +
    geom_jitter()
```

The `geom_jitter()` function allows for us to specify the amount of random
motion in the jitter, using the `width` and `height` arguments. When we don't
specify values for `width` and `height`, `geom_jitter()` defaults to 40% of the
resolution of the data (the smallest change that can be measured). Hence, if we
would like *less* spread in our jitter than was default, we should pick values
between 0.1 and 0.4. Experiment with the values to see how your plot changes.

```{r adding-width-height, fig.alt="Scatter plot of number of items owned versus number of household members, with jitter and transparency.", purl=FALSE}
interviews_plotting %>%
    ggplot(aes(x = no_membrs, y = number_items)) +
    geom_jitter(alpha = 0.5,
                width = 0.2,
                height = 0.2)
```

For our final change, we can also add colours for all the points by specifying
a `color` argument inside the `geom_jitter()` function:

```{r adding-colors, fig.alt="Scatter plot of number of items owned versus number of household members, showing points as blue.", purl=FALSE}
interviews_plotting %>%
    ggplot(aes(x = no_membrs, y = number_items)) +
    geom_jitter(alpha = 0.5,
                color = "blue",
                width = 0.2,
                height = 0.2)
```

* colour each village in the plot differently, you could use a vector as an input
to the argument **`color`**.  
* inside and outside aes 

```{r color-by-species, purl=FALSE}
interviews_plotting %>%
    ggplot(aes(x = no_membrs, y = number_items)) +
    geom_jitter(aes(color = village), alpha = 0.5, width = 0.2, height = 0.2)
```

There appears to be a positive trend between number of household
members and number of items owned (from the list provided). Additionally,
this trend does not appear to be different by village.

:::::::::::::::::::::::::::::::::::::::  challenge

## Exercise

Use what you just learned to create a scatter plot of `rooms` by `village`
with the `respondent_wall_type` showing in different colours. Does this
seem like a good way to display the relationship between these variables?
What other kinds of plots might you use to show this type of data?

:::::::::::::::  solution

## Solution

```{r scatter-challenge, fig.alt="Scatter plot showing positive trend between number of household members and number of items owned.", answer=TRUE, purl=FALSE}
interviews_plotting %>%
    ggplot(aes(x = village, y = rooms)) +
    geom_jitter(aes(color = respondent_wall_type),
	    alpha = 0.5,
		    width = 0.2,
		    height = 0.2)
```

This is not a great way to show this type of data because it is difficult to
distinguish between villages. What other plot types could help you visualize
this relationship better?

:::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::::::::::::::::

## Boxplot

We can use boxplots to visualize the distribution of rooms for each
wall type:

```{r boxplot, fig.alt="Box plot of number of rooms by wall type.", purl=FALSE}
interviews_plotting %>%
    ggplot(aes(x = respondent_wall_type, y = rooms)) +
    geom_boxplot()
```

By adding points to a boxplot, we can have a better idea of the number of
measurements and of their distribution:

```{r boxplot-with-jitter, fig.alt="Previous plot with dot plot added as additional layer to show individual values. Boxplot layer is transparent.", purl=FALSE}
interviews_plotting %>%
    ggplot(aes(x = respondent_wall_type, y = rooms)) +
    geom_boxplot(alpha = 0) +
    geom_jitter(alpha = 0.5,
    		color = "tomato",
    		width = 0.2,
    		height = 0.2)
```

We can see that muddaub houses and sunbrick houses tend to be smaller than
burntbrick houses.

Notice how the  boxplot layer is behind the jitter layer? What do you need to
change in the code to put the boxplot layer in front of the jitter layer?

:::::::::::::::::::::::::::::::::::::::  challenge

## Exercise

Boxplots are useful summaries, but hide the *shape* of the distribution. For
example, if the distribution is bimodal, we would not see it in a
boxplot. An alternative to the boxplot is the violin plot, where the shape
(of the density of points) is drawn.

- Replace the box plot with a violin plot; see `geom_violin()`.

:::::::::::::::  solution

## Solution

```{r violin-plot}
interviews_plotting %>%
  ggplot(aes(x = respondent_wall_type, y = rooms)) +
  geom_violin(alpha = 0) +
  geom_jitter(alpha = 0.5, color = "tomato")
```

:::::::::::::::::::::::::

So far, we've looked at the distribution of room number within wall type. Try
making a new plot to explore the distribution of another variable within wall
type.

- Create a boxplot for `liv_count` for each wall type. Overlay the boxplot
  layer on a jitter layer to show actual measurements.

:::::::::::::::  solution

## Solution

```{r boxplot-exercise, fig.alt="Box plot of number of livestock owned by wall type, with dot plot added as additional layer to show individual values."}
interviews_plotting %>%
   ggplot(aes(x = respondent_wall_type, y = liv_count)) +
   geom_boxplot(alpha = 0) +
   geom_jitter(alpha = 0.5, width = 0.2, height = 0.2)
```

:::::::::::::::::::::::::

- Add colour to the data points on your boxplot according to whether the
  respondent is a member of an irrigation association (`memb_assoc`).

:::::::::::::::  solution

## Solution

```{r boxplot-exercise-factor, fig.alt="Previous plot with dots colored based on whether respondent was a member of an irrigation association."}
interviews_plotting %>%
  ggplot(aes(x = respondent_wall_type, y = liv_count)) +
  geom_boxplot(alpha = 0) +
  geom_jitter(aes(color = memb_assoc), alpha = 0.5, width = 0.2, height = 0.2)
```

:::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::::::::::::::::

## Barplots

Barplots are also useful for visualizing categorical data. By default,
`geom_bar` accepts a variable for x, and plots the number of instances each
value of x (in this case, wall type) appears in the dataset.

```{r barplot-1, fig.alt="Bar plot showing counts of respondent wall types."}
interviews_plotting %>%
    ggplot(aes(x = respondent_wall_type)) +
    geom_bar()
```

We can use the `fill` aesthetic for the `geom_bar()` geom to colour bars by
the portion of each count that is from each village.

```{r barplot-stack, fig.alt="Stacked bar plot of wall types showing each village as a different color."}
interviews_plotting %>%
    ggplot(aes(x = respondent_wall_type)) +
    geom_bar(aes(fill = village))
```

This creates a stacked bar chart. These are generally more difficult to read
than side-by-side bars. We can separate the portions of the stacked bar that
correspond to each village and put them side-by-side by using the `position`
argument for `geom_bar()` and setting it to "dodge".

```{r barplot-dodge, fig.alt="Bar plot of respondent wall types with each village as a separate bar."}
interviews_plotting %>%
    ggplot(aes(x = respondent_wall_type)) +
    geom_bar(aes(fill = village), position = "dodge")
```

This is a nicer graphic, but we're more likely to be interested in the
proportion of each housing type in each village than in the actual count of
number of houses of each type (because we might have sampled different numbers
of households in each village). To compare proportions, we will first create a
new data frame (`percent_wall_type`) with a new column named "percent"
representing the percent of each house type in each village. We will remove
houses with cement walls, as there was only one in the dataset.

```{r wall-type-data}
percent_wall_type <- interviews_plotting %>%
    filter(respondent_wall_type != "cement") %>%
    count(village, respondent_wall_type) %>%
    group_by(village) %>%
    mutate(percent = (n / sum(n)) * 100) %>%
    ungroup()
```

Now we can use this new data frame to create our plot showing the
percentage of each house type in each village.

```{r barplot-wall-type, fig.alt="Side by side bar plot showing percent of respondents in each village with each wall type."}
percent_wall_type %>%
    ggplot(aes(x = village, y = percent, fill = respondent_wall_type)) +
    geom_bar(stat = "identity", position = "dodge")
```

:::::::::::::::::::::::::::::::::::::::  challenge

## Exercise

Create a bar plot showing the proportion of respondents in each
village who are or are not part of an irrigation association
(`memb_assoc`). Include only respondents who answered that question
in the calculations and plot. Which village had the lowest proportion of
respondents in an irrigation association?

:::::::::::::::  solution

## Solution

```{r barplot-memb-assoc, fig.alt="Bar plot showing percent of respondents in each village who were part of association."}
percent_memb_assoc <- interviews_plotting %>%
  filter(!is.na(memb_assoc)) %>%
  count(village, memb_assoc) %>%
  group_by(village) %>%
  mutate(percent = (n / sum(n)) * 100) %>%
  ungroup()

percent_memb_assoc %>%
   ggplot(aes(x = village, y = percent, fill = memb_assoc)) +
    geom_bar(stat = "identity", position = "dodge")
```

Ruaca had the lowest proportion of members in an irrigation association.

:::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::::::::::::::::

## Adding Labels and Titles

By default, the axes labels on a plot are determined by the name of the variable
being plotted. However, **`ggplot2`** offers lots of customization options,
like specifying the axes labels, and adding a title to the plot with
relatively few lines of code. We will add more informative x-and y-axis
labels to our plot, a more explanatory label to the legend, and a plot title.

The `labs` function takes the following arguments:

- `title` -- to produce a plot title
- `subtitle` -- to produce a plot subtitle (smaller text placed beneath the title)
- `caption` -- a caption for the plot
- `...` -- any pair of name and value for aesthetics used in the plot (e.g.,
  `x`, `y`, `fill`, `color`, `size`)

```{r barplot-wall-types-labeled, fig.alt="Previous plot with plot title and labells added."}
percent_wall_type %>%
    ggplot(aes(x = village, y = percent, fill = respondent_wall_type)) +
    geom_bar(stat = "identity", position = "dodge") +
    labs(title = "Proportion of wall type by village",
         fill = "Type of Wall in Home",
         x = "Village",
         y = "Percent")
```

## Faceting

Rather than creating a single plot with side-by-side bars for each
village, we may want to create multiple plot, where each plot shows the
data for a single village. This would be especially useful if we had
a large number of villages that we had sampled, as a large number of
side-by-side bars will become more difficult to read.

**`ggplot2`** has a special technique called *faceting* that allows the
user to split one plot into multiple plots based on a factor included
in the dataset. We will use it to split our barplot of housing type
proportion by village so that each village has its own panel in a
multi-panel plot:

```{r barplot-faceting, fig.alt="Bar plot showing percent of each wall type in each village."}
percent_wall_type %>%
    ggplot(aes(x = respondent_wall_type, y = percent)) +
    geom_bar(stat = "identity", position = "dodge") +
    labs(title="Proportion of wall type by village",
         x="Wall Type",
         y="Percent") +
    facet_wrap(~ village)
```

Click the "Zoom" button in your RStudio plots pane to view a larger
version of this plot.

Usually plots with white background look more readable when printed.  We can set
the background to white using the function `theme_bw()`. Additionally, you can remove
the grid:

```{r barplot-theme-bw, fig.alt="Bar plot showing percent of each wall type in each village, with black and white theme applied.", purl=FALSE}
percent_wall_type %>%
    ggplot(aes(x = respondent_wall_type, y = percent)) +
    geom_bar(stat = "identity", position = "dodge") +
    labs(title="Proportion of wall type by village",
         x="Wall Type",
         y="Percent") +
    facet_wrap(~ village) +
    theme_bw() +
    theme(panel.grid = element_blank())
```

What if we wanted to see the proportion of respondents in each village
who owned a particular item? We can calculate the percent of people
in each village who own each item and then create a faceted series of
bar plots where each plot is a particular item. First we need to
calculate the percentage of people in each village who own each item:

```{r percent-items-data}
percent_items <- interviews_plotting %>%
    group_by(village) %>%
    summarize(across(bicycle:no_listed_items, ~ sum(.x) / n() * 100)) %>%
    pivot_longer(bicycle:no_listed_items, names_to = "items", values_to = "percent")
```

To calculate this percentage data frame, we needed to use the `across()`
function within a `summarize()` operation. Unlike the previous example with a
single wall type variable, where each response was exactly one of the types
specified, people can (and do) own more than one item. So there are multiple
columns of data (one for each item), and the percentage calculation needs to be
repeated for each column.

Combining `summarize()` with `across()` allows us to specify first, the columns
to be summarized (`bicycle:no_listed_items`) and then the calculation. Because
our calculation is a bit more complex than is available in a built-in function,
we define a new formula:

- `~` indicates that we are defining a formula,
- `sum(.x)` gives the number of people owning that item by counting the number of `TRUE`
  values (`.x` is shorthand for the column being operated on),
- and `n()` gives the current group size.

After the `summarize()` operation, we have a table of percentages with each item
in its own column, so a `pivot_longer()` is required to transform the table into
an easier format for plotting. Using this data frame, we can now create a
multi-paneled bar plot.

```{r percent-items-barplot, fig.alt="Multi-panel bar chart showing percent  of respondents in each village and who owned each item, with no grids behid bars."}
percent_items %>%
    ggplot(aes(x = village, y = percent)) +
    geom_bar(stat = "identity", position = "dodge") +
    facet_wrap(~ items) +
    theme_bw() +
    theme(panel.grid = element_blank())
```

## **`ggplot2`** themes

In addition to `theme_bw()`, which changes the plot background to white,
**`ggplot2`** comes with several other themes which can be useful to quickly
change the look of your visualization. The complete list of themes is available
at [https://ggplot2.tidyverse.org/reference/ggtheme.html](https://ggplot2.tidyverse.org/reference/ggtheme.html). `theme_minimal()` and
`theme_light()` are popular, and `theme_void()` can be useful as a starting
point to create a new hand-crafted theme.

The
[ggthemes](https://jrnold.github.io/ggthemes/reference/index.html)
package provides a wide variety of options (including an Excel 2003 theme). The
[**`ggplot2`** extensions website](https://exts.ggplot2.tidyverse.org/) provides a list
of packages that extend the capabilities of **`ggplot2`**, including additional
themes.

:::::::::::::::::::::::::::::::::::::::  challenge

## Exercise

Experiment with at least two different themes. Build the previous plot
using each of those themes. Which do you like best?

::::::::::::::::::::::::::::::::::::::::::::::::::

## Customization

Take a look at the [**`ggplot2`** cheat
sheet](https://github.com/rstudio/cheatsheets/blob/main/data-visualization-2.1.pdf),
and think of ways you could improve the plot.

Now, let's change names of axes to something more informative than 'village' and
'percent' and add a title to the figure:

```{r ggplot-customization, purl=FALSE}
percent_items %>%
    ggplot(aes(x = village, y = percent)) +
    geom_bar(stat = "identity", position = "dodge") +
    facet_wrap(~ items) +
    labs(title = "Percent of respondents in each village who owned each item",
         x = "Village",
         y = "Percent of Respondents") +
    theme_bw()
```

The axes have more informative names, but their readability can be improved by
increasing the font size:

```{r ggplot-customization-font-size, purl=FALSE}
percent_items %>%
    ggplot(aes(x = village, y = percent)) +
    geom_bar(stat = "identity", position = "dodge") +
    facet_wrap(~ items) +
    labs(title = "Percent of respondents in each village who owned each item",
         x = "Village",
         y = "Percent of Respondents") +
    theme_bw() +
    theme(text = element_text(size = 16))
```

Note that it is also possible to change the fonts of your plots. If you are on
Windows, you may have to install the [**`extrafont`**
package](https://github.com/wch/extrafont), and follow the instructions included
in the README for this package.

After our manipulations, you may notice that the values on the x-axis are still
not properly readable. Let's change the orientation of the labels and adjust
them vertically and horizontally so they don't overlap. You can use a 90-degree
angle, or experiment to find the appropriate angle for diagonally oriented
labels. With a larger font, the title also runs off. We can add "\\n" in the string
for the title to insert a new line:

```{r ggplot-customization-label-orientation, fig.alt="Multi-panel bar charts showing percent of respondents in each village and who owned each item, with grids behind the bars.", purl=FALSE}
percent_items %>%
    ggplot(aes(x = village, y = percent)) +
    geom_bar(stat = "identity", position = "dodge") +
    facet_wrap(~ items) +
    labs(title = "Percent of respondents in each village \n who owned each item",
         x = "Village",
         y = "Percent of Respondents") +
    theme_bw() +
    theme(axis.text.x = element_text(colour = "grey20", size = 12, angle = 45,
                                     hjust = 0.5, vjust = 0.5),
          axis.text.y = element_text(colour = "grey20", size = 12),
          text = element_text(size = 16))
```

If you like the changes you created better than the default theme, you can save
them as an object to be able to easily apply them to other plots you may create.
We can also add `plot.title = element_text(hjust = 0.5)` to centre the title:

```{r ggplot-custom-themes, purl=FALSE}
grey_theme <- theme(axis.text.x = element_text(colour = "grey20", size = 12,
                                               angle = 45, hjust = 0.5,
                                               vjust = 0.5),
                    axis.text.y = element_text(colour = "grey20", size = 12),
                    text = element_text(size = 16),
                    plot.title = element_text(hjust = 0.5))


percent_items %>%
    ggplot(aes(x = village, y = percent)) +
    geom_bar(stat = "identity", position = "dodge") +
    facet_wrap(~ items) +
    labs(title = "Percent of respondents in each village \n who owned each item",
         x = "Village",
         y = "Percent of Respondents") +
    grey_theme
```

:::::::::::::::::::::::::::::::::::::::  challenge

## Exercise

With all of this information in hand, please take another five minutes to
either improve one of the plots generated in this exercise or create a
beautiful graph of your own. Use the RStudio [**`ggplot2`** cheat sheet](https://github.com/rstudio/cheatsheets/blob/main/data-visualization-2.1.pdf)
for inspiration. Here are some ideas:

- See if you can make the bars white with black outline.
- Try using a different colour palette (see
  [http://www.cookbook-r.com/Graphs/Colors\_(ggplot2)/](https://www.cookbook-r.com/Graphs/Colors_\(ggplot2\)/)).

::::::::::::::::::::::::::::::::::::::::::::::::::

After creating your plot, you can save it to a file in your favourite format. The Export tab in the **Plot** pane in RStudio will save your plots at low resolution, which will not be accepted by many journals and will not scale well for posters.

Instead, use the `ggsave()` function, which allows you to easily change the dimension and resolution of your plot by adjusting the appropriate arguments (`width`, `height` and `dpi`).

Make sure you have the `fig_output/` folder in your working directory.

```{r ggsave-example, eval=FALSE, purl=FALSE}
my_plot <- percent_items %>%
    ggplot(aes(x = village, y = percent)) +
    geom_bar(stat = "identity", position = "dodge") +
    facet_wrap(~ items) +
    labs(title = "Percent of respondents in each village \n who owned each item",
         x = "Village",
         y = "Percent of Respondents") +
    theme_bw() +
    theme(axis.text.x = element_text(color = "grey20", size = 12, angle = 45,
                                     hjust = 0.5, vjust = 0.5),
          axis.text.y = element_text(color = "grey20", size = 12),
          text = element_text(size = 16),
          plot.title = element_text(hjust = 0.5))

ggsave("fig_output/name_of_file.png", my_plot, width = 15, height = 10)
```

Note: The parameters `width` and `height` also determine the font size in the saved plot.

:::::::::::::::::::::::::::::::::::::::: keypoints

- `ggplot2` is a flexible and useful tool for creating plots in R.
- The data set and coordinate system can be defined using the `ggplot` function.
- Additional layers, including geoms, are added using the `+` operator.
- Boxplots are useful for visualizing the distribution of a continuous variable.
- Barplots are useful for visualizing categorical data.
- Faceting allows you to generate multiple plots based on a categorical variable.

::::::::::::::::::::::::::::::::::::::::::::::::::


